Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Word Processors

### 2018-07-31 main.tex
Howard updates

@@

\noindent\textbf{Findings:} We created 396 mappings and translated 58.8\% of the \gls*{dlmf} (2,405 expressions) successfully between \Maple{} and \gls*{dlmf}. For a significant percentage, the definitions in Maple and the \gls*{dlmf} were different. Consequently, an atomic symbol in one system maps to a composite expression in the other system. The translator was also successfully used for automatic verification processes of mathematical online compendia and \gls*{cas}. Our evaluation techniques discovered two errors in the \gls*{dlmf} and one bug in \Maple.

@@

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occur for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(0)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce a \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
as \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
as \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.

Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry sufficient semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

@@

Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care of each translation and provide details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

The position of the branch cut of the arccotangent function differs after translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. In consequence, translation~(\ref{eq:acot-alternatives-1}) solves the issue, as long as the user do not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z = -\iunit$.

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{ll}
		\hline
		Semantic Macro Entry & Maple Entry \\
		\hline
		\verb|\LegendreP{\nu}@{x}| & \verb|LegendreP($0, $1)| \\
		\verb|X1:\LegendrePX\LegendreP[\mu]{\nu}@{x}| & \verb|LegendreP($1, $0, $2)|\\
		\hline
	\end{tabular}
	\caption{Example entries of the Legendre and associated Legendre function in the translation library. The prefix notation \texttt{X<d>:<name>X} defines the translation for \texttt{<name>} with \texttt{<d>}-number of optional arguments.}
	\label{tab:legendreP-lex}
\end{table}

@@

\begin{wrapfigure}{r}{0.38\textwidth}
\vspace{-10pt}
\begin{minipage}{0.38\textwidth}
\center
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\vspace{-5pt}
\caption{The mathematical expression `$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{minipage}
\vspace{-5pt}
\end{wrapfigure}

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects like lists. This allows to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three example for the \gls*{teo} list that groups some tokens.

@@

As you can see, the input expression in semantic \LaTeX{} is interpreted as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^2 \right)^2.
\end{equation}
 \section{Maple to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
Instead of writing a custom \Maple{} syntax parser, we using \Maple's internal data structure to get a syntax tree of the input\footnote{In consequence, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allow different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} using \gls*{dag} as syntax trees.

@@

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls*{cas} have trouble to test even simple equations. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \textit{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \textit{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. Because of implementation details, there are some techniques that helps \Maple{} to find possible simplifications. For example we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an equivalent representation using the exponential representations
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions we are able to improve the simplification process in \Maple. However, the limitations of the \textit{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one has to use analytic continuation for the right-hand side of equation (\ref{eq:branch-cut-near}).

To evaluate such complex cases, the equivalence checks of \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

\subsection{Results}\label{sec:test-summary}
There are 685 \Macro s\footnote{The macros are still work in progress, and therefore the total number is constantly changing.} in total, and 665 of them were implemented in the translator engine. We defined forward translations to \Maple{} for 201 and backward translations from \Maple{} for 195 functions.

@@

The numerical test engine was able to discover a sign error in equation~\parencite[(14.5.14)]{NIST:DLMF}\footnote{The equation had originally been stated as shown in equation~(\ref{eq:signerror}). The error were reported on 10th April 2017.}
\begin{equation}\label{eq:signerror}
\displaystyle \mathsf{Q}^{-1/2}_{\nu}\left(\cos\theta\right)=-\left(\frac{\pi}{2\sin\theta}\right)^{1/2}\frac{\cos\left(\left(\nu+\frac{1}{2}\right)\theta\right)}{\nu+\frac{1}{2}}.
\end{equation}
The error can be found on~\parencite[p. 359]{NIST:Handbook} and has been fixed in the \gls*{dlmf} with version 1.0.16. The same engine also identified a missing comma in the constraint of~\parencite[(10.16.7)]{NIST:DLMF}. The original constraint was given by $2\nu \neq -1, -2 -3, \ldots$, with a missing comma after $-2$.

@@

The tasks for future work are diverse. The main task to improve the translator by implementing more functions and features. For example, for the current state, only translations to \Maple's standard function library were implemented. \Maple{} allows to load extra packages dynamically and therefore support several more functions. This feature would drastically increase the number of possible translations. With such improvements, further work on evaluation techniques become worthwhile to evaluate \gls*{dlmf} and \gls*{cas}.

Furthermore, the translator was designed to be easily extendable. This allows to implement translations for other \gls*{cas} without much effort.

The biggest weakness of the translator is still the semantic macros. Currently, we are working on mathematical information retrieval techniques to allow an extension for the translator for generic \LaTeX{} inputs.

>>>>

\DeclareLabeldate[online]{%
  \field{date}
  \field{year}
  \field{eventdate}
  \field{origdate}
  \field{urldate}
}

@@

\noindent\textbf{Findings:} We created 396 mappings and translated 58.8\% of the \gls*{dlmf} (2,405 expressions) successfully between \Maple{} and \gls*{dlmf}. For a significant percentage, the definitions in \Maple{} and the \gls*{dlmf} were different. Consequently, an atomic symbol in one system maps to a composite expression in the other system. The translator was also successfully used for automatic verification processes of mathematical online compendia and \gls*{cas}. Our evaluation techniques discovered two errors in the \gls*{dlmf} and one bug in \Maple.

@@

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occur for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{}\footnote{The mention of
specific products, trademarks, or brand
names is for purposes of identification only. Such mention is not to be interpreted
in any way as an endorsement or certification of such products or brands by the
National Institute of Standards and Technology, nor does it imply that the products
so identified are necessarily the best available for the purpose. All trademarks
mentioned herein belong to their respective owners.} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(0)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce a \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
as \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
as \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017.}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry sufficient semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

@@

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care of each translation and provide details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. In consequence, translation~(\ref{eq:acot-alternatives-1}) solves the issue, as long as the user do not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z = -\iunit$.

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{ll}
		\hline
		Semantic Macro Entry & \Maple{} Entry \\
		\hline
		\verb|\LegendreP{\nu}@{x}| & \verb|LegendreP($0, $1)| \\
		\verb|X1:\LegendrePX\LegendreP[\mu]{\nu}@{x}| & \verb|LegendreP($1, $0, $2)|\\
		\hline
	\end{tabular}
	\caption{Example entries of the Legendre and associated Legendre function in the translation library. The prefix notation \texttt{X<d>:<name>X} defines the translation for \texttt{<name>} with \texttt{<d>}-number of optional arguments.}
	\label{tab:legendreP-lex}
\end{table}

@@

\begin{wrapfigure}[4]{r}{0.38\textwidth}
\vspace{-10pt}
\begin{minipage}{0.38\textwidth}
\center
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\vspace{-5pt}
\caption{The mathematical expression `$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{minipage}
\vspace{-5pt}
\end{wrapfigure}

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects like lists. This allows one to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three example for the \gls*{teo} list that groups some tokens.

@@

As you can see, the input expression in semantic \LaTeX{} is interpreted as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^2 \right)^2.
\end{equation}
 \section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
Instead of writing a custom \Maple{} syntax parser, we using \Maple's internal data structure to get a syntax tree of the input\footnote{In consequence, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allow different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} using \gls*{dag} as syntax trees.

@@

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls*{cas} have trouble to test even simple equations. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \textit{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \textit{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. Because of implementation details, there are some techniques that helps \Maple{} to find possible simplifications. For example we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an equivalent representation using the exponential representations
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions we are able to improve the simplification process in \Maple. However, the limitations of the \textit{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one has to use analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the equivalence checks of \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

\subsection{Results}\label{sec:test-summary}
There are 685 \Macro{}s\footnote{The DLMF/DRMF semantic macros are still a work in progress, and the total number is constantly changing.} in total, and 665 of them were implemented in the translator engine. We defined forward translations to \Maple{} for 201 and backward translations from \Maple{} for 195 functions.

@@

The numerical test engine was able to discover a sign error in equation~\parencite[(14.5.14)]{NIST:DLMF}\footnote{The equation had originally been stated as shown in (\ref{eq:signerror}). The error was reported on 10th April 2017.}
\begin{equation}\label{eq:signerror}
\displaystyle \mathsf{Q}^{-1/2}_{\nu}\left(\cos\theta\right)=-\left(\frac{\pi}{2\sin\theta}\right)^{1/2}\frac{\cos\left(\left(\nu+\frac{1}{2}\right)\theta\right)}{\nu+\frac{1}{2}}.
\end{equation}
The error can be found on~\parencite[p. 359]{NIST:Handbook} and has been fixed in the \gls*{dlmf} with version 1.0.16. The same engine also identified a missing comma in the constraint of~\parencite[(10.16.7)]{NIST:DLMF}. The original constraint was given by $2\nu \neq -1, -2 -3, \ldots$, with a missing comma after the $-2$.

@@

The tasks for future work are diverse. The main task to improve the translator by implementing more functions and features. For example, for the current state, only translations to \Maple's standard function library were implemented. \Maple{} allows one to load extra packages dynamically and therefore support several more functions. This feature would drastically increase the number of possible translations. With such improvements, further work on evaluation techniques become worthwhile to evaluate \gls*{dlmf} and \gls*{cas}.

The translator was designed to be easily extendable. This allows one to implement translations for other \gls*{cas} without much effort.
The biggest weakness of the translator is still the semantic macros {\bf [HSC: I don't understand what this sentence means. I think you need to be more specific here. The way it is written it doesn't make sense.]}. Currently, we are working on mathematical information retrieval techniques to allow an extension for the translator for generic \LaTeX{} inputs.
